\large UNet-HCRF Integration for Sequence Labeling on EEG Data

### 2024-12-09 manuscript/slides.tex
edit methods

\begin{frame}{CNN-HCRF Model Structure}
\begin{itemize}
\item An UNet CNN was designed to generate a feature map, and a global
potential for the label.
\item A hidden conditional random field (HCRF) layer processes the pixel-wise
label information from the feature map, assigning a latent label $h_i$ to each
pixel.
\item The latent label $h_i$ contributes to the global feature through the
following potetial:
\begin{equation*}
\varphi(y, h_{j}; \delta) = \sum_{a \in Y} \sum_{b \in H} \delta_{a, b}
\cdot \one(y = a) \cdot \one(h_j = b),
\end{equation*}
where $\delta_{a,b}$ denotes the likelihood of class label $y = a$
containing a joint with hidden state $h_j = b$, learned during training.
\end{itemize}
\end{frame}

\begin{frame}{CNN-HCRF Model Structure (cont.)}
\begin{small}
\begin{itemize}
\item Unary potential $\phi(h_j, x_j; \omega)$: Measures the likelihood of the
local feature $x_j$ is assigned as the hidden part state $h_j$, whose parameter
$\omega$ is learned by end-to-end CNN-UNet structure.
\item Binary potential $\psi(h_i, h_j, x_i, x_j; \eta)$: Balances the hidden
label compatibility of the neighboring pixels $p$ on the feature image 
$\Fb(\Ib)$:
\begin{align*}
\psi(h_i, h_j, x_i, x_j; \eta) &= \mu(h_i, h_j) \Bigg[
\omega_1 \exp \left(
-\frac{\left\lvert p_i - p_j \right\rvert^2}{2\eta_\alpha^2}
-\frac{\left\lvert \Fb_i(\Ib) - \Fb_j(\Ib) \right\rvert^2}{2\eta_\beta^2}
\right) \\
&+ \omega_2 \exp \left(
- \frac{\left\lvert p_i - p_j \right\rvert^2}{2 \eta_\gamma^2}
\right)
\Bigg],
\end{align*}
where the parameters $\eta_\alpha$, $\eta_\beta$ and $\eta_\gamma$ control the
influence of the corresponding feature spaces.
\item Global potential $\vartheta(y, \xb; \varpi)$: Measures the compatibility
of class label $y$ and the global feature vector $x_0$ of the whole action,
whose parameter $\varpi$ will be learned during training process.
\end{itemize}
\end{small}
\end{frame}

\begin{frame}{CNN-HCRF Model Structure (cont..)}
Let $y$ represent the corresponding label of the input image $\xb$.
The jointly probability distribution of the HCRF model is:
\begin{align*}
\Phi(y, \hb, \xb; \theta) &= \underbrace{
\sum_{j \in \nu} \phi(h_j, x_j; \omega) +
\sum_{i \neq j} \psi(h_i, h_j, x_i, x_j; \eta)}_{
% \propto \log\PP( \hb \vert \xb; \theta)
\textrm{Measures log-likelihood $\log\PP(\hb \vert \xb; \theta)$}} \\
& + \underbrace{
\sum_{j \in \nu} \varphi(y, h_j, x_j; \delta) + \vartheta(y, \xb; \varpi)}_{
% \propto \log\PP(y | \hb, \xb; \theta)
\textrm{Measures log-likelihood $\log\PP(y \vert \hb, \xb; \theta)$}}
\end{align*}
\end{frame}

@@

\begin{frame}{Model Inference}
\begin{small}
\begin{itemize}
\item The inference for the latent variable follows a standard mean-field
Variational Inference scheme.
\begin{equation*}
\QQ(h_i) \propto \exp\{\EE_{\QQ(\hb_{-i})}
\log\PP(y, h_i, \hb_{-i} \vert \xb; \theta)\}.
\end{equation*}
\item The resulting ELBO will be used as predicted log-likelihood of each image:
\begin{equation*}
\begin{split}
& \textrm{ELBO}(\QQ) = \EE_{\QQ(\hb)} \log\PP(y, \hb \vert \xb; \theta)
- \EE_{\QQ(\hb)} \log q(\hb) \\
&\quad = \sum_{i=1}^N \EE_{q_i(h_i)}
\left[ \phi(h_j, x_j; \omega) + \varphi(y, h_j, x_j; \delta) \right] \\
&\quad + \sum_{i=1}^N \sum_{j=1}^N \sum_{l,l^\prime} \one_{\{i \neq j\}} 
\QQ(h_i = l) \QQ(h_j = l^\prime) \mu(l, l^\prime) \\
&\quad \left[ \omega_1 \exp \left(
- \frac{\left\lvert p_i - p_j \right\rvert^2}{2\eta_\alpha^2}
- \frac{\left\lvert \Fb_i(\Ib) - \Fb_j(\Ib) \right\rvert^2}{2\eta_\beta^2}
\right) + \omega_2 \exp \left(
- \frac{\left\lvert p_i - p_j \right\rvert^2}{2\eta_\gamma^2}
\right) \right] \\
&\quad - \sum_{i=1}^N \EE_{q_i} \log q_i(h_i) + \vartheta(y, \xb; 
\varpi).
\end{split}
\end{equation*}
\end{itemize}
\end{small}
\end{frame}

>>>>

\begin{frame}{CNN-HCRF Model Structure}
\begin{itemize}
\item Given the feature of the input data $\Ib$ learned by CNN, i.e.
$\xb = \Fb(\Ib)$, its corresponding hidden part states $\hb$, and the class
label $y$, a hidden conditional random field (HCRF) has the exponential form:
\begin{equation*}
\PP(y, \hb \vert \xb; \theta)
= \frac{\exp\left(\Phi(y, \hb, \xb; \theta)\right)}
{\sum_{y^\prime \in Y} \sum_{\hb \in H^N}
\exp\left(\Phi(y^\prime, \hb, \xb; \theta)\right)},
\end{equation*}
where $\theta$ is the model parameter, $H^N$ denotes the set of all possible
hidden part states of $N$ hidden parts, and $\Phi(y, \hb, \xb; \theta)$
refers to potential function depending on the feature $\xb$.
\item The probability of class label $y$ for the given feature $\xb$ is:
\begin{equation*}
\PP(y \vert \xb; \theta) = \sum_{\hb \in H^N} \PP(y, \hb \mid \xb; \theta)
= \frac{\sum_{\hb \in H^N} \exp\left(\Phi(y, \hb, \xb; \theta)\right)}
{\sum_{y^\prime \in Y} \sum_{\hb \in H^N}
\exp\left(\Phi(y^\prime, \hb, \xb; \theta)\right)}.
\end{equation*}
\end{itemize}
\end{frame}

\begin{frame}{CNN-HCRF Model Structure (cont.)}
The jointly probability distribution of the HCRF model is:
\begin{align*}
\Phi(y, \hb, \xb; \theta) &= \underbrace{
\sum_{j \in \nu} \phi(h_j, x_j; \omega) +
\sum_{i \neq j} \psi(h_i, h_j, x_i, x_j; \eta)}_{
% \propto \log\PP( \hb \vert \xb; \theta)
\textrm{Measures log-likelihood $\log\PP(\hb \vert \xb; \theta)$}} \\
& + \underbrace{
\sum_{j \in \nu} \varphi(y, h_j, x_j; \delta) + \vartheta(y, x_0; \varpi)}_{
% \propto \log\PP(y | \hb, \xb; \theta)
\textrm{Measures log-likelihood $\log\PP(y \vert \hb, \xb; \theta)$}},
\end{align*}
where:
\begin{itemize}
\item Unary potential $\phi(h_j, x_j; \omega)$: Measures the likelihood of the
local feature $x_j$ is assigned as the hidden part state $h_j$, with parameter
$\omega$ is learned by end-to-end CNN-UNet structure.
\end{itemize}
\end{frame}

\begin{frame}{CNN-HCRF Model Structure (cont..)}
\begin{itemize}
\item Binary potential $\psi(h_i, h_j, x_i, x_j; \eta)$: Balances the hidden
label compatibility of the neighboring pixels $p$ on the feature $\xb$:
\begin{align*}
\psi(h_i, h_j, x_i, x_j; \eta) &= \mu(h_i, h_j) \Bigg[
\omega_1 \exp \left(
-\frac{\left\lvert p_i - p_j \right\rvert^2}{2\eta_\alpha^2}
-\frac{\left\lvert x_i - x_j \right\rvert^2}{2\eta_\beta^2}
\right) \\
&+ \omega_2 \exp \left(
- \frac{\left\lvert p_i - p_j \right\rvert^2}{2 \eta_\gamma^2}
\right)
\Bigg],
\end{align*}
where $\mu(h_i, h_j)$ is a label compatibility function,
%that describes the interactive influences between different pairs of classes.
$\omega_1$ and $\omega_2$ are linear combination weights of different kernels,
and the parameters $\eta_\alpha$, $\eta_\beta$ and $\eta_\gamma$ control the
influence of the corresponding feature spaces.
\end{itemize}
\end{frame}

\begin{frame}{CNN-HCRF Model Structure (cont...)}
\begin{itemize}
\item Unary potential $\varphi(y, h_{j}; \delta)$: Measures the compatibility
between global class label $y$ and the hidden local state $h_j$:
\begin{equation*}
\varphi(y, h_{j}; \delta) = \sum_{a \in Y} \sum_{b \in H} \delta_{a, b}
\cdot \one(y = a) \cdot \one(h_j = b),
\end{equation*}
where $\one(\cdot)$ is a indicator function, and $\delta_{a,b}$ denotes the
likelihood of class label $y = a$ containing a joint with hidden state
$h_j = b$, learned during training.
\item Global potential $\vartheta(y, x_0; \varpi)$: Measures the likelihood of
the global feature $x_0$ is assigned as the action label $y$, with parameter
$\varpi$ learned during training process.
\end{itemize}
\end{frame}

\begin{frame}{Variational Inference of HCRF}
\begin{small}
\begin{itemize}
\item The mean-field varepsilon family approximation of
$\log P(\hb \mid y,\xb,\theta)$ is expressed as
$\QQ(h) = \prod_{i=1}^N q_i(h_i)$.
\item The evidence lower bound (ELBO) under the mean-field family is:
\begin{equation*}
\begin{split}
& \textrm{ELBO}(\QQ) = \EE_{\QQ(\hb)} \log\PP(y, \hb \vert \xb; \theta)
- \EE_{\QQ(\hb)} \log q(\hb) \\
&\quad = \sum_{i=1}^N \EE_{q_i(h_i)}
\left[ \phi(h_j, x_j; \omega) + \varphi(y, h_j, x_j; \delta) \right] \\
&\quad + \sum_{i=1}^N \sum_{j=1}^N \sum_{l,l^\prime} \one_{\{i \neq j\}} 
\QQ(h_i = l) \QQ(h_j = l^\prime) \mu(l, l^\prime) \\
&\quad \left[ \omega_1 \exp \left(
- \frac{\left\lvert p_i - p_j \right\rvert^2}{2\eta_\alpha^2}
- \frac{\left\lvert \Fb_i(\Ib) - \Fb_j(\Ib) \right\rvert^2}{2\eta_\beta^2}
\right) + \omega_2 \exp \left(
- \frac{\left\lvert p_i - p_j \right\rvert^2}{2\eta_\gamma^2}
\right) \right] \\
&\quad - \sum_{i=1}^N \EE_{q_i} \log q_i(h_i) + \vartheta(y, \xb; 
\varpi).
\end{split}
\end{equation*}
\end{itemize}
\end{small}
\end{frame}

\begin{frame}{Variational Inference of HCRF (cont.)}
\begin{itemize}
\item By the coordinate ascent variational inference(CAVI) algorithm, fixing
$q_j(h_j), \forall j \neq i$, the optimal $q_i(h_i)$ that maximizes
$\mathrm{ELBO}(q_i(h_i))$ is:
\begin{equation*}
\QQ(h_i) \propto \exp\{\EE_{\QQ(\hb_{-i})}
\log\PP(y, h_i, \hb_{-i} \mid \xb; \theta)\}.
\end{equation*}
\item The explicit form of $q_i(l)$ is derived as follows:
\begin{equation*}
\begin{split}
& q_i(l) = \frac{1}{Z_i} \exp \Bigg\{
\phi(h_j, x_j; \omega) + \varphi(y, h_j, x_j; \delta)
+ \sum_{j \neq i} \sum_{l^\prime} q(l^\prime) \mu(l, l^\prime) \\
& \left[ \omega_1 \exp \left(
- \frac{\left\lvert p_i - p_j \right\rvert^2}{2\eta_\alpha^2}
- \frac{\left\lvert \Fb_i(\Ib) - \Fb_j(\Ib) \right\rvert^2}{2\eta_\beta^2}
\right)
+ \omega_2 \exp \left(
- \frac{\left\lvert p_i - p_j 
\right\rvert^2}{2\eta_\gamma^2}
\right)
\right]
\Bigg\},
\end{split}
\end{equation*}
where $Z_i$ is the normalization constant.
\end{itemize}
\end{frame}

@@

%\begin{frame}{Model Inference}
%\begin{small}
%\begin{itemize}
%\item The inference for the latent variable follows a standard mean-field
%Variational Inference scheme.
%\begin{equation*}
%\QQ(h_i) \propto \exp\{\EE_{\QQ(\hb_{-i})}
%\log\PP(y, h_i, \hb_{-i} \vert \xb; \theta)\}.
%\end{equation*}
%\item The resulting ELBO will be used as predicted log-likelihood of each 
%image:
%\begin{equation*}
%\begin{split}
%& \textrm{ELBO}(\QQ) = \EE_{\QQ(\hb)} \log\PP(y, \hb \vert \xb; \theta)
%- \EE_{\QQ(\hb)} \log q(\hb) \\
%&\quad = \sum_{i=1}^N \EE_{q_i(h_i)}
%\left[ \phi(h_j, x_j; \omega) + \varphi(y, h_j, x_j; \delta) \right] \\
%&\quad + \sum_{i=1}^N \sum_{j=1}^N \sum_{l,l^\prime} \one_{\{i \neq j\}} 
%\QQ(h_i = l) \QQ(h_j = l^\prime) \mu(l, l^\prime) \\
%&\quad \left[ \omega_1 \exp \left(
%- \frac{\left\lvert p_i - p_j \right\rvert^2}{2\eta_\alpha^2}
%- \frac{\left\lvert \Fb_i(\Ib) - \Fb_j(\Ib) \right\rvert^2}{2\eta_\beta^2}
%\right) + \omega_2 \exp \left(
%- \frac{\left\lvert p_i - p_j \right\rvert^2}{2\eta_\gamma^2}
%\right) \right] \\
%&\quad - \sum_{i=1}^N \EE_{q_i} \log q_i(h_i) + \vartheta(y, \xb; 
%\varpi).
%\end{split}
%\end{equation*}
%\end{itemize}
%\end{small}
%\end{frame}
